
# Metadata in Open Science


Metadata is the backbone of digital curation. Without it a digital resource may be irretrievable, unidentifiable or unusable. 

- **Descriptive Metadata** *enables identification, location and retrieval of information resources by users, often including the use of controlled vocabularies for classification and indexing and links to related resources.*

- **Technical Metadata** *describes the technical processes used to produce, or required to use a digital object.*

- **Administrative Metadata** *is used to manage administrative aspects of the digital object such as intellectual property rights and acquisition.*


This usually takes the form of a structured set of elements. Different communities develop different standards which define both the **structure** and **content** of metadata.


## Ecological Metadata Language (EML) {.smaller}

The last decade has witnessed a tremendous explosion of ecological and environmental data, with the potential to greatly enhance understanding of the complexity of the biosphere. 

However, broad-scale or synthetic research is hampered by data being:

- extremely heterogeneous and complexity 

- dispersed across many separate repositories. 

<big>data are **largely unorganized and inaccessible**</big>

<br>

**Ecological Metadata Language (EML)** is a metadata standard developed by and for the ecology discipline. EML is a set of XML schema documents that allow for the **structural expression of metadata**.

<big>**Wide adoption and use of EML will create exciting new opportunities for data discovery, access, integration and synthesis.**</big>



## Install EML r package {.smaller}

There are interface software that can be used to produce EML files but 
an r package is being developed which will allows the creation of EML files in R

```{r, eval = F}
install.packages("devtools", repos = "http://cran.us.r-project.org")
```
```{r}
library("devtools")
```

```{r, eval = F}
install_github("ropensci/EML", build=FALSE, 
               dependencies=c("DEPENDS", "IMPORTS"))
```
```{r}
library("EML")

```



## An example {.smaller}

Here's a small example dataset of counts of fish recorded across different rivers

```{r, echo=FALSE}
dat <- data.frame(river = factor(c("SAC",  
                                   "SAC",   
                                   "AM")),
                  spp   = c("Oncorhynchus tshawytscha",  
                            "Oncorhynchus tshawytscha", 
                            "Oncorhynchus kisutch"),
                  stg   = ordered(c("smolt", 
                                    "parr", 
                                    "smolt"), 
                                  levels=c("parr", 
                                           "smolt")), # => parr < smolt
                  ct    = c(293L,    
                            410L,    
                            210L),
                  day   = as.Date(c("2013-09-01", 
                                    "2013-09-1", 
                                    "2013-09-02")),
                  stringsAsFactors = FALSE)
```


```{r}
dat

```
Like many real data sets, the column headings are convenient for data entry and manipulation, but **not particularly descriptive to a user not already familiar with this data**. 

More important still, they **don't let us know what units they are measured in** (or in the case of categorical / factor data like species names or life stages, **what the factor abbreviations refer to**). So let us take a moment to be more explicit:

We can store this information by creating EML files through the function ```eml_write()``` where we explicity supply variable descriptions and units (as lists) through the  **```col.defs```** and **```unit.defs```** arguments respectively


## Metadata structure {.smaller}
<big>**```col.defs```** are relatively self explanatory, but **```unit.defs```** can be a little tricky.</big>

<br>

  - For ```factors```, this is a definition of the levels involved.

  - For ```numeric``` data, units must be specified according from the link below, or custom units defined.  
    + EML standard units 
<https://knb.ecoinformatics.org/#external//emlparser/docs/eml-2.1.1/./eml-unitTypeDefinitions.html#StandardUnitDictionary>
    + Optionally you can express the precision and bounds of the unit as well.
    
  - For ```Dates``` or ```times```, the format must be specified, (e.g. YYYY or MM-DD-YY). Optionally you can also define the precision.
  
  - For ```character strings```, a definition of the kind of string can be given

## Let's explore using our example {.smaller}

```{r}
col.defs <- c("River site used for collection",
              "Species scientific name",
              "Life Stage", 
              "count of live fish in traps",
              "The day on which traps were sampled")


unit.defs <- list(
  c(SAC = "The Sacramento River",     # Factor 
    AM = "The American River"),
 "Scientific name",                   # Character string 
  c(parr = "third life stage",        # Ordered factor 
    smolt = "fourth life stage"),
  c(unit = "number", 
    precision = 1, 
    bounds = c(0, Inf)),              # Integer
  c(format = "YYYY-MM-DD",            # Date
    precision = 1))
```

## Create EML file {.smaller}

```{r, eval = F}
eml_write(dat,
          unit.defs = unit.defs,
          col.defs = col.defs,
          creator = "Anna Krystalli",
          contact = "annakrystalli@googlemail.com",
          file = "workflow/outputs/EML/EML_example.xml")

```

The file is now ready to be published to academic repositories. Here we will publish to ```figshare``` as an example

```{r, eval=FALSE}
eml_publish("workflow/outputs/EML/EML_example.xml", 
            description="Example EML file from EML",
            categories = "Ecology", 
            tags = "EML", 
            destination="figshare") 
```


## Extracting information from EML files {.smaller}

Data and metadat can also be extracted from EML files using the ```eml_get()``` function. 
First we need to load our EML file

```{r}
obj <- eml_read("~/Documents/WORK/ACCE Data management course/workflow/outputs/EML/EML_example.xml")
```
```{r, eval=FALSE}
dat <- eml_get(obj, "data.frame")
```

dat
```
```{r}
eml_get(obj, "contact")
eml_get(obj, "citation_info")

```

## More details on EML

EML data modules
<https://knb.ecoinformatics.org/#external//emlparser/docs/eml-2.1.1/./eml-attribute.html#NumberType>

Digital Curation Center : Ecological Metadata Language resources

<http://www.dcc.ac.uk/resources/metadata-standards/eml-ecological-metadata-language>