---
output: html_document
---
##**ACCE Data Management Workshop Survey Report**
####***Anna Krystalli***
#####***12 April 2015***


####**INTRODUCTION**

On the 17th February 2017, myself (Dr. Anna Krystalli) delivered the one-day workshop on data management, covering tidy data, metadata and open science in R, developed for first year NERC ACCE Doctoral Partnership students back in 2015. The workshop, attended by 25 students was carried out at the University of York and was part of a three day course on quantitative skills. Following the workshop we invited participants to complete an online feedback survey to determine how useful, challenging and likely the workshop was to change their data management practices. 13 students completed the questionnaire and we summarise the responses gathered in this report. We also include demographic characteristics of the respondents and the transcript of responses to open-ended feedback questions in the Appendix.

####**SURVEY RESULTS**

#####**Q1 How useful did you find the course overall?**

#####**Q2 How useful did you find the introduction?**

#####**Q3 How useful did you find the session on tidy data?**

#####**Q4 How useful did you find the session on documentation and metadata?**

#####**Q5 How useful did you find the session on sharing and accessing data?**



```{r, echo=FALSE, warning=F, message=F, error=FALSE}


if (!require("pacman")) install.packages("pacman")
pacman::p_load(c("sjPlot", "likert", "dplyr"), character.only = T)


dd <- read.csv("assets/data/survey1.csv", check.names = F)
#dd <- read.csv("Rmd/assets/data/survey1.csv", check.names = F)


levels = c("Not at all useful", "Quite useful", "Useful", "Very useful")
likert <- dd[,2:6] %>% lapply(
                               FUN = function(x){
                                   ordered(x, levels = levels)
                                   }) %>% data.frame() 
names(likert) <- paste("Q", 1:5, sep = "")
likert <- likert[,ncol(likert):1]
sjp.stackfrq(likert, geom.colors = "BuGn")

```



#####**Q5 How challenging did you find the practical exercises?**



```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2}
levels = c("Too easy", "About right", "Too difficult")
likert <- dd[,7, drop = F] %>% lapply(
                               FUN = function(x){
                                   ordered(x, levels = levels)
                                   }) %>% data.frame() 
names(likert) <- paste("Q", 6, sep = "")
sjp.stackfrq(likert, geom.colors = "Paired")

```


#####**Q6 How likely are you to change your data management practices following this workshop?**



```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2}
levels<- c("Unlikely","Quite likely", "Very likely")
likert <- dd[,8, drop = F] %>% lapply(
                               FUN = function(x){
                                   ordered(x, levels = levels)
                                   }) %>% data.frame() 
names(likert) <- paste("Q", 7, sep = "")
likert <- likert[,ncol(likert):1]
sjp.stackfrq(likert, geom.colors = "BuGn")
```


#####**Q7 Do you think data management should be taught to future ACCE students?**


```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2}

levels <- c("Don't know", "Yes", "No")
likert <- dd[,9, drop = F] %>% lapply(
                               FUN = function(x){
                                   ordered(x, levels = levels)
                                   }) %>% data.frame() 
names(likert) <- paste("Q", 8, sep = "")
likert <- likert[,ncol(likert):1]
sjp.stackfrq(likert, geom.colors = "Paired")

```


####**SUMMARY**

<br>

The workshop and the majority of the individual elements of the workshop appear to have been generally successful. Participants reported greater benefits when evaluating the workshop as a whole and also reported greater utility of the section on tidy data. While respondents reported overall lower degrees of usefulness on the metadata section, it was also the section that all participants got something out of. All participants reported that they were either 'quite likely' or 'very likely' to change their data management practices is an indication that the workshop has the potential to change behaviour, but that remains to be seen in practice. Most participants (85.7%) also agree that data management should be included in ACCE DDP training.

All students found the balance of lectures and practical work 'about right' and the majority (~71.4%) found the difficulty of the practical elements pitched at an appropriate level. However, 4 ACCE students, two from Sheffield and two from Liverpool, found exercises too difficult to follow. It appears that the effectiveness of the workshop is hampered by a student's lack of R skills as none of the participants reporting difficulties in following the practicals found the course as a whole 'very useful', in contrast to participants comfortable with the level of difficulty, 52% of which found the workshop 'very useful'. 

The most problematic section was the metadata section. It seemed to be polarising, with comparatively large percentage of participants scoring the section as 'Very useful' but also with the only single participant throughout the whole survey reporting they did not find it useful at all.  This year, beacuse of progress that had been made with the introduction of the R package EML, I tried to introduce a practical implementation of turning student datasets into a fully documented Ecological Metadata Language structured file, ready for repositing. This was a step too far for both the time available and the familiarity with the topic of the students. So I plan to strip it back and simplify it, 

- introduce EML so they are familiar with the type of metadata files they will need to produce to reposit a typical ecological dataset but not in great detail. Mainly so they understanding how the structure and content allows data to be searchable, alignable and shareable.
- focusing on making sure the students have a checklist and best formats
- Outline the basic metadata files and formats required to build an EML
- In the practical begin to create the basic data sheets in which students can begin to collate their metadata
-  eventually create their structured metadata file but not attempt to make it during

### FUTURE DIRECTION

#####**Q8 `r names(dd[,17, drop = F])`**


```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2}
dd.n <- dd
dd.n[dd.n == ""] <- NA
names(dd.n) <- make.names(names(dd))
ggplot(dd.n, aes(x = How.familiar.are.you.with.version.control.)) + geom_bar(fill = "steelblue") + coord_flip() + xlab("")

```

#####**Q9 `r names(dd[,22, drop = F])`**


```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2}

levels =  rev(c("sign me up", "very interested", "somewhat interested", "not interested"))
likert <- dd[,22, drop = F] %>% lapply(
                               FUN = function(x){
                                   ordered(x, levels = levels)
                                   }) %>% data.frame() 
names(likert) <- paste("Q", 9, sep = "")
sjp.stackfrq(likert, geom.colors = "Paired")
```

#####**Q10 `r names(dd[,20, drop = F])`**


```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2}
levels =  rev(c("I use it regularly", "I've created a .Rmd document", "I know what it is but never used it", "Never heard of it!"))
likert <- dd[,20, drop = F] %>% lapply(
                               FUN = function(x){
                                   ordered(x, levels = levels)
                                   }) %>% data.frame() 
names(likert) <- paste("Q", 10, sep = "")
sjp.stackfrq(likert, geom.colors = "Paired")

```

#####**Q11 `r names(dd[,18, drop = F])`**


```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2}
levels =  rev(c("sign me up", "very interested", "somewhat interested", "not interested"))
likert <- dd[,18, drop = F] %>% lapply(
                               FUN = function(x){
                                   ordered(x, levels = levels)
                                   }) %>% data.frame() 
names(likert) <- paste("Q", 11, sep = "")
sjp.stackfrq(likert, geom.colors = "Paired")

```

#####**Q12 `r names(dd[,21, drop = F])`**


```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2}

levels =  rev(c("Yes", "Maybe", "No"))
likert <- dd[,21, drop = F] %>% lapply(
                               FUN = function(x){
                                   ordered(x, levels = levels)
                                   }) %>% data.frame() 
names(likert) <- paste("Q", 12, sep = "")
sjp.stackfrq(likert, geom.colors = "Paired")
```

<br><br><br>


```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2, eval = F}

likert <- data.frame(factor(dd[dd[,7] == "Too difficult", 2], levels = unique(dd[dd[,7] == "Too difficult", 2])[2:1]))
names(likert) <- "Q9"
levels <- levels(likert[,1])
axisLabels.y <- "Too difficult"
items <- likert


sjp.stackfrq(likert)

```


```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2, eval=F}

likert <- data.frame(factor(dd[dd[,7] == "About right", 2], levels = unique(dd[dd[,7] == "About right", 2])[c(2, 1, 3, 4)]))
names(likert) <- "Q5"
levels <- levels(likert[,1])
axisLabels.y <- "About right"
items <- likert


sjp.stackfrq(likert)

```

A simple way to address this, echoed in the comments of a number of participants, would be to place the data management workshop after the two day quantitative skills workshop.  Ultimately, students’ abilities in R are always going to vary and it will always be hard to cover all levels equally. A few students commented that they had little prior experience and/or that they were not planning to use it throughout their postgraduate studies and therefore felt there was too much focus on applying practices in R. More emphasis on the benefits of learning some basic script-based data management and analysis skills may therefore be useful in future. While there always may be a very small minority of ACCE students for which R analysis skills are not required to complete their PhD, we believe they should be encouraged to both understand why even basic R skills are an important academic asset and to engage with training. If the data management workshop and indeed the quantitative statistics course is to remain a compulsory part of ACCE DDP training perhaps a very basic introductory course to R should also be a standard  prerequisite. 


When asked to comment more generally on the course, students seemed generally positive. For example one student commented that 'it was one of the most useful and well delivered courses they had the pleasure of attending' and another described the subject of the workshop as 'a potential dry subject delivered very well'. A number of respondents described instructors as excellent/very good and demonstrating good knowledge of the subject. Respondents also commended the stucture of the course, suggesting there was something for everyone, even for participants with substantial prior programming experience.  Overall therefore we feel that the course was successful in delivering the proposed objectives and engaging participants, and appears a useful contribution to the ACCE DDP development program.

<br><br><br><br>



####**APPENDIX**

<br>

####***STUDENT DEMOGRAPHICS***

#####**Q10 Please give your institution**

```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2, eval = F}


levels <- unique(dd[,12])
likert <- dd[,12, drop = F] %>% lapply(
                               FUN = function(x){
                                   factor(x, levels = levels)
                                   }) %>% data.frame()
names(likert) <- "Q10"
sjp.stackfrq(likert, geom.colors = "Paired")

```

<br>

#####**Q12 Are you an ACCE student?**

```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2}

levels <- unique(dd[,13])
likert <- dd[,13, drop = F] %>% lapply(
                               FUN = function(x){
                                   factor(x, levels = levels)
                                   }) %>% data.frame()
names(likert) <- "Q12"

sjp.stackfrq(likert, geom.colors = "Paired")

```

<br><br>

####***RESPONSES TO OPEN-ENDED QUESTIONS***

**Note:** A number of respondents made comments on aspects of the three day course as a whole.

<br>

#####**Q8 Do you have any further general comments about the course, the instructors, or the technical level at which it was pitched? **

```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2}
library(knitr)
dat <- dd[,10]
dat <- dat[dat != ""]
kable(data.frame(no = 1:length(dat), responses = dat), type = "markdown")
```

<br>

#####**Q13 Which topics, if any, do you think we covered in too much detail? **

```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2}
library(knitr)
dat <- dd[,15]
dat <- dat[dat != ""]
dat <- unique(dat[dat != "none"])
kable(data.frame(no = 1:length(dat), responses = dat), type = "markdown")
```

<br>

#####**Q14 Are there any topics we should have covered, or covered in more detail? **

```{r, echo=FALSE, warning=F, message=F, error=FALSE, fig.height=1.2}
library(knitr)
dat <- dd[,16]
dat <- dat[dat != ""]
kable(data.frame(no = 1:length(dat), responses = dat), type = "markdown")
```