---
title: "Open Science in R"
author: "ACCE Data Management Course"
date: "10 March 2015"
output: ioslides_presentation
---

## The age of Open Science 

Data standards are allowing the compilation of disparate datasets into large openly accessible databases.

In turn, R packages are being developed to allow easy access of such resources through R

We'll explore this functionality by focusing on the **National Biodiversity Network (NBN)** database

## rNBN {.smaller}

THe NBN is an information network of biodiversity data provided by a collaboration of organisations, such as the UK nature agencies, wildlife trusts, the Natural History Museum, The Marine Biological Association etc. <http://www.nbn.org.uk/>



```rNBN``` is a statistical package developed for retrieving data from the NBN Gateway

The package utilises the NBN login facilities, ensuring that when you use it you have access to all of the data that you would have access to if you were to login to the NBN Gateway itself.

<br>

Functions within the package allow users to search occurrence data by

    - species 
    - year-range
    - location
    - dataset 
    - species groups
 
 <br>
 

## rNBN {.smaller}

First install and load the ```rnbn``` package

```{r, eval=F}
install.packages("rnbn", repos = "http://cran.us.r-project.org")

library(rnbn)
```

Have a look at the package functionality
```{r, eval =FALSE}
vignette('rnbn_vignette')

```

<file:///Users/Anna/Library/R/3.1/library/rnbn/doc/rnbn_vignette.html>

<br>

Next we need to register with NBN
<https://data.nbn.org.uk/User/Register>

and activate account 

## Get Taxon code {.smaller}

So let's try and extract some occurence data for a species of interest, the copepod *Calanus finmarchicus*

<br>

The first thing we need to do is identify the correct taxonVersionKey. We use the function **```getTVKQuery()```** to identify it.


<br>

NB the first time you call any function within the package you will be asked to authenticate. You can also use function **`nbnLogin(username = 'usr', password = 'pass')`** to login


## Get Taxon code {.smaller}

```{r, echo = 3:4, eval=T}
library(rnbn)
nbnLogin(username = 'akrystalli', password = 'JoyZipper-14')


```
```{r}
query.res <- getTVKQuery("Calanus finmarchicus")
head(query.res)
```
## Get Taxon Version Key {.smaller}

Bit confusing so let's have a closer look at the first row (transposed into a column for easier viewing)
```{r}
t(query.res[1,])
```

## Get Taxon code {.smaller}

The first row of the query result table is usually the most relevant entry and the prefered taxonVersionKey is listed under the column ```ptaxonVersionKey``` so I store the key in an object named ```cfin.TVK``` to use for quering the database.

```{r}
cfin.TVK <- query.res[1, "ptaxonVersionKey"]
cfin.TVK
```

## Get Taxon Occurences {.smaller}

I use the taxonVersionKey (```cfin.TVK```) and the function ```getOccurences()``` to extract occurences for the species of interest from the NBN database
```{r}
occ <- getOccurrences(tvks = cfin.TVK, acceptTandC = T, silent = T)

occ[1:10,c("pTaxonName", "startDate",
           "latitude", "longitude")]
```

## Plot on a map {.smaller}

I use the function ```map``` in package ```maps``` to plot the occurences extracted from the NBN database

```{r, eval = F}
install.packages("maps", repos = "http://cran.us.r-project.org")
```

```{r, eval =FALSE}
library(maps)
map(xlim = c(-10,4), ylim = range(occ$latitude), 
    fill = T, col = "black", mar = c(1,1,1,1))
points(occ$longitude, occ$latitude, col = "red", pch = "*")

```

## Plot on a map {.smaller}
```{r, echo = F}
library(maps, quietly = T)
map(xlim = c(-10,4), ylim = range(occ$latitude), 
    fill = T, col = "black", mar = c(1,1,1,1))
points(occ$longitude, occ$latitude, col = "red", pch = "*", cex = 2)

```


## Have a go! {.smaller}

  - Install rNBN package
  - Register with NBN online
    + confirm registration
  - Pick a species and try and extract the occurences
  - Have a go at mapping

<br>

Once installed details of using the package can be accessed through:
```{r, eval = F}
vignette('rnbn_vignette')
```


## Other rOpenSci packages {.smaller}

http://ropensci.org/packages/
