---
title: "TidyR - Tidying data"
author: "ACCE Data Management Course"
date: 
output: html_document
---

<br>

## Principles of tidy data {.smaller}
<br>

<big>**tidying**: structuring datasets to facilitate analysis.</big>
<br>

  - <small>**One variable per column**.</small>
  - <small>**One row per observation**.</small>

<br>

### **Why?**
+ <small>**Easier to index**.</small>
+ <small>**Easier to analyse**.</small>


*Best to record data in tidy format but ```tidyR``` package provides functions to tidy untidy data*

<br>

## Examples of messy data {.smaller}

- Column headers are values, not variable names.

- Multiple variables are stored in one column.

- Variables are stored in both rows and columns.

- Multiple types of observational units are stored in the same row.


<br>

<big> Most messy datasets can be tidied with a small set of tools: 

**gathering**, **separating** and **spreading**. </big>

<br> 

***

## **`tidyR`** {.smaller}

#### Install tidyR

```{r, message=F}
#install.packages("tidyr", "dplyr")

library(tidyr)
library(dplyr)

```

-if you are asked about storing to a personal library just type 'y' (yes)

<br> 

***

## 1 - **gathering** (wide -> long) {.smaller}
### one variable over many columns

In this experiment three (rather unusually named) people were given two different drugs (a and b) and their heart rate was recorded:

```{r}

messy <- data.frame(
  name = c("Wilbur", "Petunia", "Gregory"),
  a = c(67, 80, 64),
  b = c(56, 90, 50)
)
head(messy)

```

**How many variables have we got?**

**A sensible model we might want to fit:**

```heart rate ~ drug```


How can we supply the data to a modelling function? (e.g. ```lm()```)

<br>

We use the function ```gather()``` in the tidy package to reshape the dataframe from wide to long format


```gather(data, key, value, ..., na.rm = FALSE, convert = FALSE)```

- ```data``` :       **a data frame**
- ```key``` :   **name for the identifier of the columns to gather** 
- ```value```:  **name for the new variable being created**
- ```...```:  **select the columns to be stacked**
    + <small>you can use ```:``` to select all variables between two columns.</small> 
    + <small>use bare variable names (ie " " not required).</small>

```{r}

tidy  <- gather(data = messy, key = drug, value = heartrate, a:b)



```



```{r}
tidy
```

```heart rate ~ drug```

## 2 - **separate** (split)  {.smaller}
### more than one variable in a single column

In this study, parasite counts were made on samples taken from 3 elephant faecal boluses and two locations 


```{r}

messy <- data.frame(
  id = rep(1:2, each = 6),
  spl.id = paste(rep(rep(1:3, each = 2),2), c("centre", "outer"), sep = "-"),
  counts = rnbinom(12, mu = 2, size = 15)
)
head(messy, 8)
```




We use ```separate()``` to split ```spl.id``` into ```bolus``` and ```location```, using a regular expression to describe the character that separates them.

```separate(data, col, into, sep , remove = TRUE, convert = FALSE, extra = "error", ...)```

- ```data``` :         **a data frame**
- ```col``` :   **name of column to split** 
- ```into``` :   **name for the new variable being created** (as character vector)
- ```sep```:  character as regular expression to match to any **non-alphanumeric** values
    + numeric indicates position of split
    + **length of ```sep``` should be one less than length of ```into```**





```{r}

separate(data = messy, into = c("bolus", "location"), col = spl.id, sep= "\\-")

```


## 3 - Combination {.smaller}

In this study, the time people spent on their phones was measured at two locations (work and home) and at two time points.

```{r}

messy <- data.frame(
  id = 1:4,
  trt = sample(rep(c('control', 'treatment'), each = 2)),
  work.T1 = runif(4),
  home.T1 = runif(4),
  work.T2 = runif(4),
  home.T2 = runif(4)
)
head(messy, 8)
```



First we first use ```gather()``` to turn columns ```work.T1```, ```home.T1```, ```work.T2``` and ```home.T2``` into a key-value pair of key and time. (Only the first eight rows are shown to save space.)

```{r}

tidier <- gather(data = messy, key = key, value = time, -id, -trt)
head(tidier, 8)
```



Next we use ```separate()``` to split the key into ```location``` and ```time```, using a regular expression to describe the character that separates them.


```{r}

tidy <- separate(data = tidier, col = key, into = c("location", "timepoint"), sep = "\\.") 

head(tidy, 8)

```

<br>

***


# TidyR Data exercises {.bigger}

<br>

## Example data 1 {.smaller}

In an investigation of sexual dimorphism in a tropical bird species, wing length of male and female birds within breeding pairs were recorded across a number of regions within the species range. Reshape the data so that differences in average wing length between sexes could be tested.


[**data**](https://drive.google.com/open?id=0B9BMeZ9H2nRxZlFURnBRZlRRWEE)

<br>


```{r , echo = 2:3}
birds <- read.csv("~/Google Drive/ACCE RDM workshop/Tidy Data exercise/birds.csv")
head(birds, 20)

```


```{r}
tidy.birds <- gather(data = birds, key = sex, value = length, male:female)
head(tidy.birds, 20)
```

<br>

***

## Example data 2 {.smaller}

In an investigation of how different tree species affect tropical rainforest microclimate, measurements of temperature and humidity were made at ground level (below) and above the canopy (above) for individual trees.  



**data**

<https://drive.google.com/open?id=0B9BMeZ9H2nRxU01LWFJzeFYydXc>

```{r , echo = 2:3}
trees <- read.csv("~/Google Drive/ACCE RDM workshop/Tidy Data exercise/trees.csv")
head(trees, 20)

```



```{r}
tidy.trees <- separate(data = trees, col = spl, into = c("tree_id", "elevation"), 
                       sep = "/")
head(tidy.trees, 20)
```

<br>

***


## Example data 3 {.smaller}

In this study, daily trends in 3 different stocks (X,Y,Z) was monitored daily. We need the data reshaped in an appropriate format that would allow the trend of stocks to be examined as a function of stock identity, month or day. Beware of the date format!



[**data**](https://drive.google.com/open?id=0B9BMeZ9H2nRxdTdOa3Z5NkxxcG8)




```{r , echo = 2:3}
stocks <- read.csv("~/Google Drive/ACCE RDM workshop/Tidy Data exercise/stocks.csv")
head(stocks, 15)

```


```{r}
stocks %>% 
  gather(key = stock, value = trend, -date) %>% 
  separate(col = date, into = c( "month", "day","year"), sep = c("-", "-")) %>%  
  head(15)
```



## Further Reading

Hadley Wickam's paper on Tidy Data

<http://www.jstatsoft.org/v59/i10/paper>

RStudio Data wrangling [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

